# Data Analysis of Docking
---
## Intro
Hello everyone, this is Meiers here. I have created this document to demo a .rmd files (which lack some of the functionalities of [Jupyter Notebooks](https://jupyter.org/try-jupyter/lab/) such as being compatible with python packages) but Bucknell seems to not have Notebooks installed on any of its computers, so here we are.

As I am sure you have noticed by now, this file is written in markdown. Markdown is a way to allow for enhanced formatting of plaintext documents. RMD stands for R Markdown, in other words, it is markdown but allows for interspersed chunks of R. This is similar to jupyter notebook, but instead of python, we use R, widely accepted as a superior programming language. Jokes aside, R is a very powerful tool to visualize data which has access to libraries with unparalleled control over graph creation like ggplot and easy-to-access statistical tests.

### Data Information
In this demo, we would like to look at the binding affinity of a variety of molecules on a variety of different proteins with similar structures. So far, we have:

1. Found the most stable structure of each protein sample (output: .pdb)
2. Preformed energy minimization on this structure (output: .pdbqt)
3. Preformed docking experiments with about 200 ligands for each of our structures. (output: .pdbqt)

Now we are left with several thousand .pdbqt files, which need cleaning and then analysis

## Cleaning
To clean the data, first we must take all of our csv files and then add them to a dataframe. All of my csv files are in a subdirectory called pdbqt_files. First though, lets import packages.
```{r}
#import packages
options(warn = -1)
library(readr)
library(tidyverse)
library(progress)
library(dplyr)
```

It will be much easier to construct plots and mutate our data if we can compile it all into a single data frame. For our data, we are interested in the name, xyz coordinate positions for the ligand, as well as the affinity each ligand has. Lets construct the dataframe like so. It will start entirely empty.
```{r}
#Construct an empty dataframe, and initalize the columns
data_frame <- data.frame(
    id = numeric(),
    name = character(),
    receptor = character(),
    ligand = character(),
    x_coord = numeric(),
    y_coord = numeric(),
    z_coord = numeric(),
    affinity = numeric(),
    model = numeric(),
    stringsAsFactors = FALSE
)
```

Next we need to harvest the names all of the .pdbqt files in our pdbqt_files directory. Let's add them to a character vector called pdbqt_files.
```{r}
# Define the ligands to include
include_ligands <- c(
  "001", "002", "003", "004", "005", "010", "016", "017", "018", "019", "020",
  "021", "022", "023", "025", "026", "028", "030", "031", "034", "035", "037",
  "038", "039", "043", "044", "046", "047", "048", "049", "050", "052", "054",
  "055", "057", "058", "064", "066", "067", "068", "069", "070", "071", "073",
  "074", "075", "076", "077", "079", "080", "082", "083", "084", "085", "087",
  "089", "090", "091", "092", "093", "094", "095", "098", "100", "101", "102",
  "103", "105", "108", "109", "112", "113", "114", "116", "117", "118", "119",
  "120", "121", "122", "123", "124", "125", "126", "129", "130", "131", "133",
  "134", "135", "136", "139", "140", "141", "144", "145", "146", "147", "148",
  "149", "150", "151", "152", "153", "154", "155", "156", "157", "158", "159",
  "160", "161", "162", "164", "165", "166", "167", "168", "169", "170", "171",
  "172", "174", "176", "177", "178", "179", "182", "183", "184", "185", "186",
  "187", "188", "189", "190", "191", "193", "194", "195", "196", "197"
)

# List all pdbqt files
pdbqt_files <- list.files(path = "docking_11_4/", pattern = "\\.pdbqt$", full.names = TRUE)

# Filter files to include only those with specified ligands
# pdbqt_files <- pdbqt_files[sapply(pdbqt_files, function(file) {
#   any(sapply(include_ligands, function(ligand) grepl(ligand, basename(file))))
# })]
```

### Creating Functions
Let's first make a function that will extract the coordinates from a data file. pdbqt files have multiple models, but we will only be skimming th data of model 1.
```{r}
################################################################################
# gets the coordinates of a file
# @param file   character vector where each element is a line in a .pdbqt file
# @param line   specifies the line in file to scan, specifically a HETATM line
# @return       returns a list of coords for current model, as a string
################################################################################

get_coords <- function(file, line) {
    start_of_coords <- 32
    coord_list <- list()
    
    x_coord <- substr(file[line], start_of_coords, start_of_coords+6)
    y_coord <- substr(file[line], start_of_coords+8, start_of_coords+14)
    z_coord <- substr(file[line], start_of_coords+16, start_of_coords+22)
    coord_list <- c(coord_list, x_coord, y_coord, z_coord)
}
```

Let's now make a function that will tell us the affinity of a ligand in a pdbqt file.
```{r}
###################################################################################
# gets the affinity of a file
# @param file   character vector where each element is a line in a .pdbqt file
# @param model  int specifying which model to look through for the affinity
# @return       returns the affinity for model, as a string
###################################################################################
get_affinity <- function(file, model) {
    total_lines <- length(file)
    for (i in 1:total_lines) {
        if (substr(file[i], 1, 7) == paste0("MODEL ", as.character(model))){
            return (substr(file[i+1], 26, 29))
        }
    }
    return("error, unable to find the current model's affinity")
}
```

Finally, let's make a method that will return a list of ints for each hetatm in the first model of any give pdbqt file.
```{r}
#############################################################################################################
# gets the lines in a file which contain HETATM coords
# @param file   character vector where each element is a line in a .pdbqt file
# @param model  int which specifies which model to parse in the .pdbqt file
# @return       returns a list of ints, where each int represents a line in model which has HETATM coords
#############################################################################################################
get_hetatm_lines <- function(file, model) {
    total_lines <- length(file)
    list_hetatm_positions = list()
    in_recording_region = FALSE
    for (i in 1:total_lines) {
        if (substr(file[i], 1, 7) == paste0("MODEL ", as.character(model))){
            in_recording_region = TRUE
        }
        if (substr(file[i], 1, 7) == paste0("MODEL ", as.character(model+1))){
            in_recording_region = FALSE
            break;
        }
        if (in_recording_region)
        {   
            if (substr(file[i], 1, 6) == "HETATM" || substr(file[i], 1, 4) == "ATOM"){
                list_hetatm_positions <- c(list_hetatm_positions, i)
            }
        }
    }
    return (list_hetatm_positions)
}
```

### Creating the data table
Now we can begin the data extraction. We can iterate through all names in our pdbqt_files vector, and for each one add a new row to data_frame.
```{r}


# Initialize the progress bar
total_length <- length(pdbqt_files)
pb <- progress_bar$new(total = total_length * 9)

temp_id <- 1

for (i in 1:total_length) {  # For each pdbqt file
  for (curr_model in 1:9) {  # For each model
    
    # Open and read the file
    file <- readLines(file(pdbqt_files[i], open = "r"))
    
    # Get the affinity for the current model
    curr_affinity <- get_affinity(file, curr_model)
    
    # Get all lines with HETATM coordinates for the current model
    hetatm_lines <- get_hetatm_lines(file, curr_model)
    
    # Create vectors to accumulate coordinates
    x_coords <- c()
    y_coords <- c()
    z_coords <- c()
    
    # For each line with HETATM coordinates, extract and accumulate coordinates
    for (line_num in hetatm_lines) {
      coords <- get_coords(file, line_num)
      x_coords <- c(x_coords, as.numeric(coords[1]))
      y_coords <- c(y_coords, as.numeric(coords[2]))
      z_coords <- c(z_coords, as.numeric(coords[3]))
    }
    
    # Calculate average coordinates if we have any
    if (length(x_coords) > 0) {
      avg_x <- mean(x_coords, na.rm = TRUE)
      avg_y <- mean(y_coords, na.rm = TRUE)
      avg_z <- mean(z_coords, na.rm = TRUE)
      
      # Add a new row to data_frame with averaged coordinates
      new_row <- data.frame(
        id = temp_id,
        name = substr(pdbqt_files[i], 13, nchar(pdbqt_files[i]) - 6),
        receptor = substr(pdbqt_files[i], 13, 20),
        ligand = substr(pdbqt_files[i], 22, 24),
        avg_x_coord = avg_x,
        avg_y_coord = avg_y,
        avg_z_coord = avg_z,
        model = curr_model,
        affinity = as.numeric(curr_affinity),
        stringsAsFactors = FALSE
      )
      
      data_frame <- rbind(data_frame, new_row)
      temp_id <- temp_id + 1
    }
    
    # Update progress bar
    pb$tick()
  }
}
```

### Examing and save the data table
Awesome! We now have 26,088 rows of atom coordinates
```{r}
glimpse(data_frame)
write.csv(data_frame, "cleaned_data_tables\\starting_data_11_4.csv")
```

Now that we have a new CSV, lets make a new RMD script that will scan our csv so we can avoid making a new CSV everytime we run the next steps.