# Averaging coords
Alright, it would be nice if we could simplify our data table such that we only have the average of all the ligand atom positions rather than the coords of every single atom. Lets make a new data frame for this purpose

```{r}
#Construct an empty dataframe, and initalize the columns
data_frame_ave <- data.frame(
    id = numeric(),
    name = character(),
    receptor = character(),
    ligand = character(),
    x_coord_ave = numeric(),
    y_coord_ave = numeric(),
    z_coord_ave = numeric(),
    model = numeric(),
    affinity = numeric(),
    stringsAsFactors = FALSE
)
```

Okay, now that we have this dataframe, lets create a loop that will iterate through our starting_data.csv and average all coords with the same name

First lets install packages:
```{r}
library(tidyverse)
library(progress)
library(dplyr)
library(cowplot)
library(plotly)
```

Lets read our starting_data.csv to a new data frame.

Fantastic, now lets filter this into a new data frame that only contains models that are within the bounding box
```{r}
x_lower <- 153.0
x_upper <- 167.0
y_lower <- 177.5
y_upper <- 188.5
z_lower <- 148.5
z_upper <- 167.0

data_frame_in_box <- read.csv("cleaned_data_tables/starting_data.csv") %>%
  mutate(
    in_range = ifelse(
    avg_x_coord > x_lower & avg_x_coord < x_upper &
    avg_y_coord > y_lower & avg_y_coord < y_upper &
    avg_z_coord > z_lower & avg_z_coord < z_upper,
    1, 0)
  )
    
# glimpse(data_frame_in_box) # There are now only 1510 rows
```

I am curious to know about the distribution of the average atom positions in space, I expect the distribution to be multimodal. Lets add a new column to the data frame which measues the distance between each average atom position and the center of the box.
```{r}
find_dist <- function(x1, y1, z1, x2, y2, z2) {
  sqrt((x1 - x2)^2 + (y1 - y2)^2 + (z1 - z2)^2)
}

box_coord_x <- mean(c(x_lower, x_upper))
box_coord_y <- mean(c(y_lower, y_upper))
box_coord_z <- mean(c(z_lower, z_upper))

data_frame_in_box <- data_frame_in_box %>%
    rowwise() %>%
    mutate(distance_from_center = find_dist(
        avg_x_coord, avg_y_coord, avg_z_coord,
        box_coord_x, box_coord_y, box_coord_z
    )) %>%
    ungroup()

# ggplot(data_frame_in_box, aes(x=distance_from_center)) + 
#      geom_histogram( binwidth=0.2, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
#   theme_cowplot()

glimpse(data_frame_in_box)
```

Okay, lets try another method of visualize
```{r}
# Create the 3D scatter plot
plot <- data_frame_in_box %>%
  plot_ly(
      x = ~avg_x_coord, 
      y = ~avg_y_coord,
      z = ~avg_z_coord,
      color = ~as.factor(in_range),
      colors = c("blue", "red")
    ) %>%
  add_markers() %>%
  layout(scene = list(
    xaxis = list(title = 'X'),
    yaxis = list(title = 'Y'),
    zaxis = list(title = 'Z')
  ))

plot
```



